---
title: "Reactome Analysis Result"
date: "`r format(Sys.time(), '%d %B, %Y')`"
params:
  result: "None"
output:
  pdf_document:
    toc: false
    toc_depth: 2
    number_sections: false
    includes:
      in_header: "preamble.tex"
    keep_tex: false
    latex_engine: xelatex
subparagraph: yes
classoption: twocolumn
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, error = FALSE, fig.pos = 'H', fig.align = "center")
library(ReactomeGSA)
suppressPackageStartupMessages( library(ggplot2) )
library(cowplot)

min_p <- 0.05

result <- params$result

if (class(result) == "character") {
  result <- readRDS("padog_result.rds")
}

pathway_result <- pathways(result, p = min_p)
n_datasets <- length(names(result))

my_colors <- c("#00A0B0", "#6A4A3C", "#CC333F", "#EB6841", "#EDC951")
my_colors_light <- c("#A1E5ED", "#C49D8C", "#EB9EA5", "#F8AE97", "#F6E4A8")

# theme for all figures
my_theme <- function(base_size = 12, base_family = "", flip_axis = F) {
    flip_axis_theme <- theme(axis.text.x = element_text(angle = 45, hjust = 1))
    
    res <- theme_bw(base_size=base_size, base_family=base_family) %+replace% 
    theme(strip.background = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.border = element_blank(), 
          axis.line = element_line(colour = "#333333")
    )
    
    if (flip_axis) {
        res <- res + flip_axis_theme
    }
    
    return(res)
}
```

# Introduction

This report summarizes the pathway analysis result created by the **Reactome Analysis Service**. For more information, visit
https://www.reactome.org.

In this reports, pathways found differentially expressed at an adjusted p-values <= `r cat(min_p)` are considered to be significantly regulated.

# Analysis Overview

Number of datasets analysed: **`r length(names(result))`**

```{r, results="asis"}
for (dataset_name in names(result)) {
  cat("**", dataset_name, ":**\n", sep = "")
  cat("\n  * ", nrow(get_result(result, "pathways", dataset_name)), " pathways\n", sep = "")
  
  if ("fold_changes" %in% result_types(result)) {
    cat("\n  * ", nrow(get_result(result, "fold_changes", dataset_name)), " fold changes for genes / proteins\n", sep = "")
  }
  
  cat("\n")
}
```

```{r significant_pathways_text, results="asis"}
if (n_datasets > 1) {
  cat("Number of datasets the different pathways were found to be significantly regulated in:\n\n")
}
```

```{r significant_pathways, fig.height=2, fig.width=4, fig.align='center'}
if (n_datasets > 1) {
  # count the number of pathways significant for the different dataset combinations
  significant_in = rowSums(pathway_result[, grep("sig\\.", colnames(pathway_result))])
  plot_data <- data.frame(
    pathway = names(significant_in),
    sig_datasets = as.numeric(significant_in)
  )
  
  ggplot(plot_data[plot_data$sig_datasets > 0, ], aes(x = sig_datasets)) +
    geom_bar(fill = my_colors[1]) +
    scale_x_continuous(breaks = 1:max(plot_data$sig_datasets, na.rm = T)) +
    labs(x = "Datasets with significant regulation", y = "Number of pathways") + 
    my_theme()
}
```

# Pathway analysis

```{r figure_sizes, include=FALSE}
# number of rows and columns for a 4-panelled figure
n_4_cols <- n_datasets %% 4
if (n_4_cols == 0) n_4_cols <- 4
n_4_rows <- ceiling(n_datasets / 4)

# number of rows and columns for a 3-panelled figure
n_3_cols <- n_datasets %% 3
if (n_3_cols == 0) n_3_cols <- 3
n_3_rows <- ceiling(n_datasets / 3)

# number of rows and columns for a 2-panelled figure
n_2_cols <- n_datasets %% 2
if (n_2_cols == 0) n_2_cols <- 2
n_2_rows <- ceiling(n_datasets / 2)
```

```{r pathway_volcano, fig.width=3, fig.height=3}
library(ggplot2)
plot_objs <- list()

for (dataset_name in names(result)) {
  plot_obj <- plot_volcano(result, dataset = dataset_name)
  print(plot_obj)
}
```

These volcano plots summarize the pathway results for every dataset. Every point is one pathway. The x-axis represents the average fold-change of all genes / proteins within that pathway. The y-axis represents the p-value where "higher" values are more significant (-log10 transformation). The red line represents p = 0.05, and the blue line p = 0.01.

```{r pathway_correlation_test, results="asis"}
if (n_datasets > 1) {
  cat("## Pathway correlation\n\n")
  cat("Correlation between all datasets (if more than one). Every point represents one pathway, with the x-axis showing the average fold-change of one dataset and the y-axis the average fold-change of the other dataset. Only pathways that were observed in both datasets are shown. Points are colored based on the lowest observed significance.\n\n")
} else {
  cat("\\notebox{Since this result only contains a single datasets, some analyses are not included in this report.}\n\n")
}
```

```{r pathway_correlation, fig.width=5, fig.height=3}
if (n_datasets > 1) {
  plot_objs <- plot_correlations(result)
  
  for (plot_obj in plot_objs) {
    print(plot_obj + my_theme())
  }
}
```

\onecolumn

```{r direction_of_regulation, results="asis"}

# add function to extract a specific field across all datasets
pathway_field_string <- function(pathway, field, do_round = F, sig_bold = T, hide_low_p = T) {
  if (!pathway %in% rownames(pathway_result)) {
    stop("Error: No information available for ", pathway)
  }
  
  regulations <- c()
  
  for (dataset_name in names(result)) {
    value <- as.character( pathway_result[pathway, paste0(field, dataset_name)] )
    
    # check if the value is defined at all
    if (is.na(value)) {
      regulations <- c(regulations, "")
      next
    }
    
    if (do_round) {
      value <- round(as.numeric(value), 3)
      
      if (value < 0.001 && hide_low_p) {
        value <- "< 0.001"
      }
    }
    
    # make bold if significant
    if (sig_bold && pathway_result[pathway, paste0("sig.", dataset_name)]) {
      value <- paste0("\\textbf{", value, "}")
    }
  
    regulations <- c(regulations, value)
  }
  
  return(paste(regulations, collapse = " & "))
}

# create a table with the details of every pathway in the list
create_pathway_table <- function(pathways) {
  n_datasets <- length(names(result))

  for (pathway in pathways) {
    cat("\\textbf{", pathway_result[pathway, "Name"], " (", pathway, ")} \n", sep = "")
    
    cat("\\begin{tabular}{ l", rep(" | c", n_datasets), " }\n")
    cat(" & ", paste(gsub("_", "\\\\_", names(result)), collapse = " & "), "\\\\\n")
    cat("\\hline\n")
    cat("Regulation & ", pathway_field_string(pathway, "Direction."), "\\\\\n")
    cat("FDR & ", pathway_field_string(pathway, "FDR.", T), "\\\\\n")
    cat("Av. FC & ", pathway_field_string(pathway, "av_foldchange.", do_round = T, hide_low_p = F, sig_bold = F), "\\\\\n")
    cat("N Genes & ", pathway_field_string(pathway, "NGenes.", F, F), "\\\\\n")
    
    
    cat("\\end{tabular}\n\n")
    cat("\\textcolor{myyellow}{\\rule{\\columnwidth}{0.3pt}}\n\n")
    
    cat("\n")
  }
}

# get the list of top pathways / separately for up- and down-regulated ones
top_n <- 20
top_pathways <- c()

for (dataset_name in names(result)) {
  direction_col <- paste0("Direction.", dataset_name)
  fdr_col <- paste0("FDR.", dataset_name)
  
  pathways_up <- pathway_result[pathway_result[, direction_col] == "Up" & !is.na(pathway_result[, direction_col]), ]
  pathways_up <- pathways_up[order(pathways_up[, fdr_col], decreasing = F), ]
  pathways_up <- pathways_up[pathways_up[, fdr_col] <= min_p, ]
  
  pathways_down <- pathway_result[pathway_result[, direction_col] == "Down" & !is.na(pathway_result[, direction_col]), ]
  pathways_down <- pathways_down[order(pathways_down[, fdr_col], decreasing = F), ]
  pathways_down <- pathways_down[pathways_down[, fdr_col] <= min_p, ]
  
  if (nrow(pathways_up) > 0) {
    pathways_to_add <- ifelse(nrow(pathways_up) > top_n, top_n, nrow(pathways_up))
    top_pathways <- c(top_pathways, rownames(pathways_up)[1:pathways_to_add])  
  }
  
  if (nrow(pathways_down) > 0) {
    pathways_to_add <- ifelse(nrow(pathways_down) > top_n, top_n, nrow(pathways_down))
    top_pathways <- c(top_pathways, rownames(pathways_down)[1:pathways_to_add])  
  }
}

top_pathways <- unique(top_pathways)

# test if the regulation is the same or not
has_same_regulation <- sapply(top_pathways, function(pathway) {
  dataset_names <- names(result)
  is_sig <- as.logical( pathway_result[pathway, paste0("sig.", dataset_names)] )
  direction <- as.character( pathway_result[pathway, paste0("Direction.", dataset_names)] )
  
  sig_direction <- direction[is_sig]
  
  # check all show the same direction
  return(length(unique(sig_direction)) == 1)
})

# increase the line-spacing for tables
cat("\\renewcommand{\\arraystretch}{1.4}\n")

if (n_datasets > 1) {
  if (sum(has_same_regulation) > 0) {
    cat("# Similar regulated pathways\n\n")
    
    cat("Pathways that show the same direction on all of the datasets. The direction of regulation is only assessed for significantly regulated pathways.\n\n")
    
    create_pathway_table(top_pathways[has_same_regulation])
  }
  
  if (sum(!has_same_regulation) > 0) {
    cat("# Differently regulated pathways\n\n")
    
    create_pathway_table(top_pathways[!has_same_regulation])
  }
} else {
  cat("# Top-regulated pathways\n\n")
  create_pathway_table(top_pathways)
}
```

# Protein / Gene level results

Classical differential expression analysis performed on the genes / proteins of every dataset. 

**Note:** Depending on the gene set analysis method used, the approach used to assess differential expression at the gene / protein level may vary to the approach used
for the pathway level.

```{r volcano_plots, fig.width=6, fig.height=n_2_rows * 3}
# create volcano plots for every dataset
plot_data <- data.frame()

for (dataset_name in names(result)) {
  dataset_fc <- get_result(result, "fold_changes", dataset_name)
  dataset_fc$dataset <- dataset_name
  
  plot_data <- rbind(plot_data, dataset_fc)
}

library(ggplot2)
ggplot(plot_data, aes(x = logFC, y = -log10(adj.P.Val))) +
  geom_point() +
  facet_wrap(~ dataset, ncol=2, scales = "free") +
  geom_hline(yintercept = -log10(0.05), color = my_colors[3], linetype = 2) +
  geom_hline(yintercept = -log10(0.01), color = my_colors[1], linetype = 2) +
  my_theme()
```

## Top up-regulated protein/genes

```{r top_up_genes, results="asis"}

for (dataset_name in names(result)) {
  cat("### ", dataset_name, "\n\n", sep = "")
  dataset_fc <- get_result(result, "fold_changes", dataset_name)
  dataset_up <- dataset_fc[dataset_fc$logFC > 0, ]
  dataset_up <- dataset_up[order(dataset_up$adj.P.Val, decreasing = F), ]
  
  print(knitr::kable(dataset_up[1:10, ], row.names = F))
  
  cat("\n\n")
}

```

## Top down-regulated protein/genes

```{r top_down_genes, results="asis"}

for (dataset_name in names(result)) {
  cat("### ", dataset_name, "\n\n", sep = "")
  dataset_fc <- get_result(result, "fold_changes", dataset_name)
  dataset_up <- dataset_fc[dataset_fc$logFC < 0, ]
  dataset_up <- dataset_up[order(dataset_up$adj.P.Val, decreasing = F), ]
  
  print(knitr::kable(dataset_up[1:10, ], row.names = F))
  
  cat("\n\n")
}

```
