---
title: "Reactome Analysis Result"
date: "`r format(Sys.time(), '%d %B, %Y')`"
params:
  result: "None"
  include_disease: "NA"
  include_interactors: "NA"
output:
  pdf_document:
    toc: false
    toc_depth: 2
    number_sections: false
    includes:
      in_header: "preamble.tex"
    keep_tex: false
    latex_engine: xelatex
subparagraph: yes
classoption: onecolumn
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, error = FALSE, fig.pos = 'H', fig.align = "center")
library(ReactomeGSA)
suppressPackageStartupMessages( library(ggplot2) )
library(cowplot)

min_p <- 0.05

result <- params$result

if (class(result) == "character") {
  result <- readRDS("/home/jg/Projects/reactome/ReactomeGSA.report/gsva_result.rds")
}

pathway_result <- pathways(result, p = min_p)
n_datasets <- length(names(result))

my_colors <- c("#00A0B0", "#6A4A3C", "#CC333F", "#EB6841", "#EDC951")
my_colors_light <- c("#A1E5ED", "#C49D8C", "#EB9EA5", "#F8AE97", "#F6E4A8")

# theme for all figures
my_theme <- function(base_size = 12, base_family = "", flip_axis = F) {
    flip_axis_theme <- theme(axis.text.x = element_text(angle = 45, hjust = 1))
    
    res <- theme_bw(base_size=base_size, base_family=base_family) %+replace% 
    theme(strip.background = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.border = element_blank(), 
          axis.line = element_line(colour = "#333333")
    )
    
    if (flip_axis) {
        res <- res + flip_axis_theme
    }
    
    return(res)
}
```

# Introduction

This report summarizes the pathway analysis result created by the **Reactome Analysis Service**. For more information, visit
https://www.reactome.org.

This report summarizes the result of a gene set variation analysis (currently the "ssGSEA" method).

# Analysis Overview

```{r basic_info, results="asis"}
# Basic information about the settings used

cat("  * Number of datasets analysed: ", length(names(result)), "\n", sep = "")
cat("  * Reactome version: ", result@reactome_release, "\n", sep = "")

if (!is.na(include_disease)) {
  disease_string = ifelse(include_disease, "", "**not**")
  cat("  * Disease pathways were ", disease_string, " included\n", sep = "")
}

if (!is.na(include_interactors)) {
  interactor_string = ifelse(include_interactors, "", "**not**")
  cat("  * Protein-protein interactors were ", interactor_string, " added\n", sep = "")
}
```

## Dataset summaries

```{r basic_dataset_info, results="asis"}
for (dataset_name in names(result)) {
  cat("**", dataset_name, ":**\n", sep = "")
  cat("\n  * ", nrow(get_result(result, "pathways", dataset_name)), " pathways\n", sep = "")
}

cat("\n")
```

# Pathway overview

```{r}
top_pathways <- 30
```

These plot's only show the top-`r top_pathways` most different pathways in every dataset.

```{r pathway_heatmaps, results="asis"}
for (dataset_name in names(result)) {
  pathway_expression <- get_result(result, "pathways", dataset_name)
  
  # find the maximum differently expressed pathway
  max_difference <- do.call(rbind, apply(pathway_expression, 1, function(row) {
      values <- as.numeric(row[3:length(row)])
      return(data.frame(name = row[1], min = min(values), max = max(values)))
  }))
  
  max_difference$diff <- max_difference$max - max_difference$min
  
  # sort based on the difference
  max_difference <- max_difference[order(max_difference$diff, decreasing = T), ]
  
  top_pathway_ids <- max_difference[1:top_pathways, "name"]
  
  cat("## Top pathways in ", dataset_name, "\n")
  
  plot_gsva_heatmap(object = result, pathway_ids = top_pathway_ids, truncate_names = T, 
                    dendrogram = "none", Rowv = F, scale = "row", key = F, 
                    rowsep = 1:length(top_pathway_ids), colsep = 1:13, sepcolor = "black", sepwidth = c(0.01, 0.0001),
                    lwid=c(0.1,4), lhei = c(0.1, 4))
}
```

# Pathway expression

These plots display the pathway-level expression values for the `r top_pathways` most different pathways per dataset.

```{r top_pathway_expressions, results="asis", fig.height=4}
for (dataset_name in names(result)) {
  cat("## Top pathways in ", dataset_name, "\n")
  
  for (pathway_id in top_pathway_ids) {
    plot_obj <- ReactomeGSA::plot_gsva_pathway(result, pathway_id)
    print(plot_obj)
  }
}
```
